---
title: "Exercise 1 - Solution"
number-sections: true
format:
  html:
    toc: true
    toc-location: left
    theme: default
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# RCT or A/B Test

## Data Simulation
```{r}
set.seed(1234)
n <- 10000
treatment <- rbinom(n, 1, 0.5)
outcome <- rbinom(n, 1, 0.3) * treatment + rbinom(n, 1, 0.5) * (1 - treatment)
```

## Results

- $P(O=1|T=1) - P(O=1|T=0)$ = `r mean(outcome[treatment==1]) - mean(outcome[treatment==0])`
- $adjustment(1) - adjustment(0)$ doesn't apply
- ATE = 0.3 - 0.5 = -0.2
- $\hat{ATE} = P(O=1|do(T=1)) - P(O=1|do(T=0))$ = `r mean(outcome[treatment==1]) - mean(outcome[treatment==0])`

# Adjustment formula

```{r}
adjustment <- function(t, o, z, t0) {
    ind_t0 <- t == t0
    z_values <- unique(z)

    adjusted_prob <- 0
    for (z_ in z_values) {
        ind_z_ <- z == z_
        ind <- ind_t0 & ind_z_
        adjusted_prob <- adjusted_prob + mean(o[ind]) * mean(ind_z_)
    }

    return(adjusted_prob)
}
```


# Confounder

## Data Simulation
```{r}
set.seed(1234)
n <- 10000

confounder <- rbinom(n, 1, 0.8)
treatment <- rbinom(n, 1, 0.6) * confounder + rbinom(n, 1, 0.2) * (1 - confounder)
outcome <- rbinom(n, 1, 0.3) * treatment + rbinom(n, 1, 0.5) * (1 - treatment) + confounder + rnorm(n)
```

## Results

- $P(O=1|T=1) - P(O=1|T=0)$ = `r mean(outcome[treatment==1]) - mean(outcome[treatment==0])`
- $adjustment(1) - adjustment(0)$ = `r adjustment(treatment, outcome, confounder, 1) - adjustment(treatment, outcome, confounder, 0)`
- ATE = 0.3 - 0.5 = -0.2
- $\hat{ATE} = P(O=1|do(T=1)) - P(O=1|do(T=0))$ = `r adjustment(treatment, outcome, confounder, 1) - adjustment(treatment, outcome, confounder, 0)`

# Mediator

## Data Simulation
```{r}
set.seed(1234)
n <- 10000

treatment <- rbinom(n, 1, 0.4)
mediator <- rbinom(n, 1, 0.6) * treatment + rbinom(n, 1, 0.2) * (1 - treatment)
outcome <- rbinom(n, 1, 0.4) * treatment + rbinom(n, 1, 0.5) * (1 - treatment) + mediator + rnorm(n)
```

## Results

- $P(O=1|T=1) - P(O=1|T=0)$ = `r mean(outcome[treatment==1]) - mean(outcome[treatment==0])`
- $adjustment(1) - adjustment(0)$= `r adjustment(treatment, outcome, mediator, 1) - adjustment(treatment, outcome, mediator, 0)`
- ATE = (0.6 + 0.4) - (0.2 + 0.5) = 0.3
- $\hat{ATE} = P(O=1|do(T=1)) - P(O=1|do(T=0))$ = `r mean(outcome[treatment==1]) - mean(outcome[treatment==0])`

# Predictor

## Data Simulation
```{r}
set.seed(1234)
n <- 10000

treatment <- rbinom(n, 1, 0.4)
predictor <- rbinom(n, 1, 0.4)
outcome <- rbinom(n, 1, 0.4) * treatment + rbinom(n, 1, 0.5) * (1 - treatment) + predictor + rnorm(n)

```

## Results

- $P(O=1|T=1) - P(O=1|T=0)$ = `r mean(outcome[treatment==1]) - mean(outcome[treatment==0])`
- $adjustment(1) - adjustment(0)$= `r adjustment(treatment, outcome, predictor, 1) - adjustment(treatment, outcome, predictor, 0)`
- ATE = 0.4 - 0.5 = -0.1

$\hat{ATE}$ has two unbiased estimators

- $\hat{ATE} = P(O=1|do(T=1)) - P(O=1|do(T=0))$ = `r mean(outcome[treatment==1]) - mean(outcome[treatment==0])`
- $\hat{ATE} = P(O=1|do(T=1)) - P(O=1|do(T=0))$ = `r adjustment(treatment, outcome, predictor, 1) - adjustment(treatment, outcome, predictor, 0)`

# Unobserved Confounder

## Data Simulation
```{r}
set.seed(1234)
n <- 10000

confounder <- rbinom(n, 1, 0.8)
treatment <- rbinom(n, 1, 0.6) * confounder + rbinom(n, 1, 0.2) * (1 - confounder)
outcome <- rbinom(n, 1, 0.3) * treatment + rbinom(n, 1, 0.5) * (1 - treatment) + confounder + rnorm(n)

```

## Results

- $P(O=1|T=1) - P(O=1|T=0)$ = `r mean(outcome[treatment==1]) - mean(outcome[treatment==0])`
- $adjustment(1) - adjustment(0)$ cannot be calculated
- ATE = -0.2
- $\hat{ATE} = P(O=1|do(T=1)) - P(O=1|do(T=0))$ cannot be calculated

# Part II

$P^{do(C:=2)}(E) = P(E|C=2) \sim N(2, 1)$

$P^{do(E:=2)}(C) \sim N(0, 1) ≠ P(C|E=2) = P(N1| N1+N2=2) \sim N(1, (1/2)1/2)$

