---
title: "Chains, forks and colliders"
number-sections: true
format:
  html:
    toc: true
    toc-location: left
    theme: default
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```



# Chains

```{r}
set.seed(1234)

n <- 10000

x <- rbinom(n, 1, 0.5)
y <- x * rbinom(n, 1, 0.3) + (1-x)*rbinom(n, 1, 0.7)
z <- y * rbinom(n, 1, 0.1) + (1-y)*rbinom(n, 1, 0.9)

cor(x, z)
```

The correlation between x and z, conditioning on y = 1 or y = 0, is close to zero, while the unconditional correlation between x and z it is not.
```{r}
print(cor(x[y == 1], z[y == 1]))
print(cor(x[y == 0], z[y == 0]))
```


# Forks

```{r}
z <- rbinom(n, 1, 0.5)
y <- z * rbinom(n, 1, 0.3) + (1-z)*rbinom(n, 1, 0.7)
x <- z * rbinom(n, 1, 0.3) + (1-z)*rbinom(n, 1, 0.7)

cor(y, x)
```

The correlation between x and y, conditioning on z = 1 or z = 0, is close to zero, while the unconditional correlation between x and y it is not.
```{r}
print(cor(x[z == 1], y[z == 1]))
print(cor(x[z == 0], y[z == 0]))
```


# Colliders

```{r}
x <- rbinom(n, 1, 0.5)
y <- rbinom(n, 1, 0.5)
z <- (1-x*y) * rbinom(n, 1, 0.2) + x*y*rbinom(n, 1, 0.8)

cor(y, x)
```

The correlation between x and y is close to zero, while conditioning on z = 1 or z = 0 it is not.

```{r}
print(cor(x[z == 1], y[z == 1]))
print(cor(x[z == 0], y[z == 0]))
```

