---
title: "Predictive efficiency"
number-sections: true
format:
  html:
    toc: true
    toc-location: left
    theme: default
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

# Data simulation: a predictor of the outcome

```{r}
set.seed(1234)

n <- 100

z <- rnorm(n, sd=.5)
x <- rnorm(n, sd=.5)
y <- 3*x + 3*z + rnorm(n, sd=.5)
```

# Controling for a predictor of the outcome can reduce ATE's standard error 

```{r}
summary(lm(y ~ x))
```

```{r}
summary(lm(y ~ x + z))
```

The standard error (variance of the estimation) of the second model is much lower than the first one. In this case, adding a covariate improves our estimation. 


#  Data simulation: a predictor of the treatment variable


```{r}
z <- rnorm(n, sd=.5)
x <- z + rnorm(n, sd=.1)
y <- 3*x + rnorm(n, sd=.5)
```

# Controling for a predictor of the treatment increases ATE's standard error

```{r}
summary(lm(y ~ x))
```

# Obtaining a linear model of bonuses and ghosts

```{r}
summary(lm(y ~ x + z))
```

The standard error for the coefficient of 'x' is higher when controling for 'z'